Effect of the Nominal Value of Foreign Currencies in valuing a good

### 2024-04-17 paper.tex
final draft

@@

\date{2024-04-16}
\begin{document}
\maketitle
\begin{abstract}
Research into the effect of using foreign currencies on consumption
behaviour suggests that consumers systematically under- or overspend
based on the nominal (face) value of the foreign currency. However,
previous research is undecided on which direction this effect has, with
the two most prominent models leading to different and mutually
exclusive outcomes. This paper considers the different models and tests
their predictive and explanatory accuracy using a study. It finds that
individuals tend to underspend when the foreign currency is less
numerous than (a fraction of) the home currency and overspend when the
opposite is the case.
\end{abstract}

@@

Past experience suggests that we treat foreign currencies differently to
our home currency and have difficulty adjusting to them. Previous
research, such as the ``money illusion'' effect suggests that
individuals are biased towards the nominal (or face value) of prices.
Past research has investigated whether a systemic effect

RAS (Raghubir and Srivastava 2002) examined the effect of using foreign
currencies on consumption behaviour and argue that there is a systematic
difference in spending behaviour based on the exchange rate of the
foreign currency to an individuals' home currency. Specifically, they
showed that when a foreign currency is a fraction of the home currency
(e.g.~1 unit in home currency is 0.2 units in foreign currency)
consumers tend to overspend in real terms relative to their home
currency. On the other hand, when the foreign currency is a multiple of
the home currency (e.g.~1 unit in home currency is 5 units in foreign
currency) consumers tend to underspend in real terms.\footnote{For
  consistency, I will use the terms fraction and multiple currencies
  throughout this paper as defined here.} RAS argue that this effect
arises because individuals are anchored on the face value of the foreign
currency which influences the exchange rate calculations from and to
their home currencies. Hence, RAS term the result of this inadequate
adjustment the face value effect.

WER (Wertenbroch, Soman, and Chattopadhyay 2007) consider the findings
by RAS and reiterate their conclusion that individuals inadequately
calculate the values of goods between differnet currencies, leading to
systematic differences in spending behaviour based on the exchange rate
between the foreign and home currencies. However, WER argue that the
process of adjustment occurs differently, namely that individuals
calculate their valuations of goods based on salient reference points,
such as budget constraints. Contrary to standard theory, WER argue that
consumers care about the difference between a valuation and their budget
rather than the ratio of the two. In the context of foreign currencies,
the nominal value of this difference will diverge, leading to systematic
differences in spending behaviour. However, the direction of this bias
is the exact opposite of the one shown by RAS. Specifically, WER show
that consumers underspend when the foreign currency is a fraction of the
home currency and overspend in the opposite case.

Fang (Fang 2019) further builds on top of the research by RAS and aims
to extend it to the domain of ``virtual'' currencies, which refers to
non-fiat currencies that are digital in nature and not controlled and
issued by a central bank. Examples of virtual currencies include video
game (or in-game) currencies, rewards points, airmiles, cryptocurrencies
and more to the extent that these are at least somewhat liquid.
Investigating in-game currencies specifically, Fang shows that consumer
valuations systematically deviate when using these currencies relative
to the home currency. Fang's findings correspond with those of RAS and
the face value effect.

This paper aims to build on the aforementioned research in several
important ways. Foremost, previous research is conclusive that there is
an effect that using foreign currencies has on individuals' consumption
behaviour relative to using their home currencies, but is divided on the
direction of this effect. Hence, this paper attempts to evaluate the
different models posed and use a study to test which model most
adequately predicts the perceived behaviour. Second, previous research
has only assigned minor significance to using incentive-compatible
mechanisms to elicit true preferences of participants. These mechanisms
have been established to elicit more accurate responses from
participants that are closer to their true preferences (\textbf{cite?}).
Hence, this study will use the Becker-DeGroot-Marshak (BDM) mechanism to
elicit participants' willingness to pay (WTP).{[}\^{} The BDM mechanism
will be defined and explained in more detail in the Methodology section
below.{]} RAS (Raghubir and Srivastava 2002) expressed the concern that
``despite attempts to make the tasks as realistic as possible, the
studies reported were laboratory experiments, and issues of
generalizability when real money is on the line do arise. For instance,
when people actually exchange foreign currency they may feel richer or
poorer depending on the exchange rate, and the differences in
perceptions of wealth may affect product valuation''. Similarly, Fang
(Fang 2019) noted that implementing the BDM mechanism was infeasible for
the study he conducted but that future research should focus on using
something like it to elicit WTP. Finally, this paper aims to extend the
previous literature on the effects of using virtual currencies
specifically, due to their increasing rise in importance when
transacting or making purchasing decisions online.

First, I present the theoretical frameworks that have been outlined by
previous research in a way that makes it easy to compare and contrast
them. These frameworks are the standard model described in consumer
choice theory, the model suggested by RAS and the model suggested by
WER. Then, I consider what each of these frameworks would predict in a
concrete case and show that all three models come to different and
mutually exclusive conclusions. In order to test which of these
frameworks is most accurate in predicting observed behaviour, I conduct
a study in which two groups each formulate their WTP for a specific good
in an incentive compatible way, with the only difference between the
groups being the exchange rate of the currency used to their home
currency and, as a result, the face values of their budget and bid.
Analysing the study, I find that consumers tend to underspend when the
foreign currency is a fraction and overspend when the foreign currency
is a multiple, which is predicted by the module proposed by WER.
However, due to the limited sample size of the study, this result is not
statistically significant and does not have significant power. Finally,
in the discussion, I will evaluate the implications of the findings for
the theoretical frameworks outlined, consider alternative explanations
for the perceived effect and argue that the model proposed by WER is the
most accurate in predicting and describing consumer behaviour when using
foreign currencies.

@@

Standard consumer choice theory suggests that individuals formulate
valuations of goods in relation to reference points, such as their
budget (\textbf{cite?}). Specifically, standard theory poses that
consumers evaluate their valuation for a good as a ratio with the budget
in order to determine what proportion of their spending power is
required to purchase the good. To formaulate valuations of goods in a
different currency, consumers would first determine the value of the
good in their home currency and subsequently apply the exchange rate to
get their valuation in the foreign currency. Hence, the model follows a
two step process:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Formaulate valuation of good as ratio to budget in home currency
\item
  Convert the valuation into foreign currency
\end{enumerate}

@@

where V is the valuation in the foreign currency, \(V_n\) is the
valuation of the good in the home currency (nominal), \(V_r\) is the
valuation of the good in the foreign currency, \(M_n\) is the budget in
the home currency, X is the exchange rate and \(f(x)\) is some function
that uses a valuation relative to the budget in order to compute
\(V_n\).

One assumption that is implicit in this model is that a consumer will
always correctly adjust their valuation ratio based on the exact
exchange rate. However, this might not always be the most rational thing
to do. For example, when exchange rates fluctuate it might not be worth
the effort of a consumer to always use the most up-to-date exchange
rate, but rather to use some approximate value. Further, when an
individual needs to do these exchange rate calculations in their head,
it might not be worth the mental effort to calculate the amount in the
foreign currency perfectly, but an approximation might be fine. For
example, it would be rational to multiply an amount in home currency by
1.2 if the real exchange rate is 1.2113248 and the effort required is
significant, such as when attempting to calculate it in one's hand. By
loosening these assumptions, the standard model is able to account for
small deviations from a consumer's real valuation of a good in a foreign
currency.

@@

RAS (Raghubir and Srivastava 2002) put forth a different model for how
consumers formulate their valuation of a good in a foreign currency
based on their studies conducted. Specifically, they pose that a
consumer will first formulate a valuation of a good in their home
currency, before converting that amount into the foreign currency using
the exchange rate. However, they argue that this calculation is biased
by the nominal value of the good in the home currency (in cases where a
consumer does not determine their valuation but observes a price, this
calculation would be biased by the nominal value of the good in the
foreign currency). This model may be formulated as follows:

@@

where V is the valuation in the foreign currency, \(V_n\) is the
valuation of the good in the home currency (nominal), \(V_r\) is the
valuation of the good in the foreign currency and α is a weighting
parameter such that \(α ∈[0,1]\). Importantly, RAS (Raghubir and
Srivastava 2002) argue that \(α > 0\) and that V is hence not equal to
\(V_r\) but is biased towards \(V_n\).

@@

WER (Wertenbroch, Soman, and Chattopadhyay 2007) build on top of RAS,
but put forth a different model that they argue is more representative
of the behaviour of actual consumers. Specifically, they agree with the
standard model that valuations are made in relation to salient reference
points, such as a consumers' budget, and argue that the model put forth
by RAS fails to account for this. However, they disagree with the
standard model that valuations use a ratio of price to budget, but
instead argue that consumers think in terms of differences. Hence,
instead of considering what fraction of their purchasing power is
required to buy a good, consumers consider how much of their budget is
left over. However, WER argue that consumers inadequately adjust this
difference to the exchange rate and are thus biased when considering
goods in different currencies since the nominal differences between
price and budget will differ when considering different currencies. This
model can be formulated as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Formaulate valuation of good as difference to budget in home currency
\item
  Convert the valuation into foreign currency
\item
  Adjust using the difference to budget in the foreign currency
\end{enumerate}

@@

where V is the valuation in the foreign currency, \(V_n\) is the
valuation of the good in the home currency (nominal), \(V_r\) is the
valuation of the good in the foreign currency, \(M_r\) is the budget in
the foreign currency (real) and α is a weighting parameter such that
\(α ∈[0,1]\). Importantly, WER (Wertenbroch, Soman, and Chattopadhyay
2007) argue that \(α > 0\) and that V is hence not equal to \(V_r\) but
is biased towards \(M_r - V_r\).

@@

where \(V_r\) is the valuation of the good in the foreign currency.

@@

The study itself was split into four parts: Consent Form, Introduction,
Auction and Follow-Up Questions. The entire experiment instructions can
be found in the Experiment Instructions section of the Appendix. The
first section asked participants to read through and sign the consent
form in order to ensure that they were informed about the study and what
data would be collected. In the introductory section, the BDM mechanism
was explained to participants and they were asked to complete four
question that tested their understanding of the mechanism. The aim of
this section was to ensure that participants properly understood how the
mechanism worked in order to elicit their real WTP during the main part
of the study.

The auction was the main part of the study, with the aim of eliciting
the participants' WTP on a specific item and comparing these across
different currencies used. In this section, participants were randomly
split into two groups with almost-identical instructions. The
participants were instructed to bid on the item shown below, a tabletop
airhockey table, using the virtual currency ``BlueCoin''. The auction
was explained to follow the BDM mechanism and it was also stated that
three participants of the experiment would be randomly selected, their
auction would be simulated and they would receive the outcome. The only
difference between the groups was the exchange rate between BlueCoin and
the British Pound (£) and the endowment that the participants received
in BlueCoin. The first group received an endowment of 10 BlueCoin where
1 BlueCoin equals £5 and the second group received an endowment of 250
BlueCoin where 1 BlueCoin equals £0.2. To complete the auction,
participants were asked to respond with their bid, in BlueCoin.

@@

Since the aim of the study was to determine whether there were any
systematic differences in WTP between consumers when the only differing
factor is the currency used, we decided to not include a control group
that bid on the item using Pounds, which was the design of some of the
experiments run by RAS (Raghubir and Srivastava 2002). This allowed for
a simpler design and more power given the same sample size. Further,
unlike the studies run by RAS (Raghubir and Srivastava 2002) and WER
(Wertenbroch, Soman, and Chattopadhyay 2007), we did not use a fiat
currency or mix of fiat and fictional currencies, but used only one
fictional currency for both groups. On the one hand, this simplification
was due to a focus of wether the effects described in previous papers
also held in similar ways for virtual currencies. On the other hand, it
also allowed us to preclude any biases or previous experience that
participants might have had towards or with certain currencies. RAS
(Raghubir and Srivastava 2002) acknowledge that ``the face value effect
is due to the accessibility and perceptual salience of the face value of
the foreign currency \ldots{} {[}which{]} is likely to depend on the
extent to which an individual has the opportunity or the time available
to process exchange rate information and/or has experience in using a
particular foreign currency''. To bias participants the least, we chose
the relatively nondescript name for a fictional, virtual currency:
BlueCoin.

Another important decision in the design of the experiment was which
item to select that a majority of participants would have a non-zero
WTP. This is important since if the majority or even the entirety of
participants were to bid zero on the item, then we would only learn that
this was the participants' WTP for that good, but not whether there are
any systematic differences in the WTP when using different currencies.
Hence, we selected a tabletop airhockey table, something that few enough
people have so that they won't want another one, but enough people want
with a non-zero WTP, especially among a student population. In fact,
only six participants of the experiment stated a WTP of 0.

In designing the main part of the experiment, we chose to conduct an
auction, specifically using the BDM mechanism, in order to have an
incentive-compatible method of eliciting the participants' WTP. The
BDM's incentive compatibility in eliciting WTP is a well-established
phenomenon that is superior to using participants' stated preferences
(\textbf{cite?}). As mentioned above, the studies conducted by RAS
(Raghubir and Srivastava 2002), WER (Wertenbroch, Soman, and
Chattopadhyay 2007) (with the exception of one permutation of one of the
studies) and Fang (Fang 2019) did not implement incentive-compatible
methods of eliciting the participants' preferences, making this
experiment and important addition to these previous studies in examining
whether the effects describe still hold when a more accurate way of
eliciting WTP is used. In order to incentivise the participants even
further, we ran a lottery that paid out three randomly-chosen
participants based on their actual choices made in the auction and by
simulating the BDM mechanism using a random number generator. Due to
budget constraints, we were only able to play this lottery for three
people, but given that the sample size of the experiment was relatively
small, the potential to receive one's outcome is likely to have helped
in eliciting the participants' true WTP.

Finally, given our budget, we were able to give participants an
endowment of £50 each, which was around double of what the item costs to
buy. Assuming that the item is priced roughly around the average WTP
(which turned out to be correct), this endowment was chosen to leave
enough room for people with higher WTP to make their bid and for us to
capture this in the data. This is important for the goals of the
experiment, since if we had, for example, chosen an endowment of £20,
the majority of participants might have bid their maximum amount,
hindering our ability to meaningfully compare WTP across the two groups.
We chose relatively simple exchange rates of 5 and \(1/5\) so that the
exchange rate calculations would be relatively easy to make, allowing us
to be more confident that differences between groups were not due to the
mental effort required to do calculations but due to something else. RAS
(Raghubir and Srivastava 2002) acknowledge that ``the reliance on face
value may be a function of the ease with which the foreign money can be
converted. \ldots{} Future research should examine the asymmetric and
nonlinear nature of this effect''. Since the goal of this study is to
investigate that an effect exists that is not solely based on a high
difficulty or large mental effort required when doing specific,
difficult calculations, we aimed to make the calculations as easy as
possible in order to isolate the effect of the exchange rate and nominal
value of the foreign currency.

@@

In setting up the experiment, we incorrectly used the BDM mechanism.
Instead of explaining it as a sealed-bid second-price auction, the
experiment instructions explained it as a sealed-bid first-price
auction. Unlike in a second-price auction, the participants of a first
price auction do not have a dominant strategy to bid their true
valuations, in fact there generally is no dominant strategy in such an
auction (\textbf{cite?}). (\textbf{todo?})

>>>>

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{241,243,245}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}

@@

\date{2024-04-16}
\begin{document}
\maketitle
\begin{abstract}
Research into the effect of using foreign currencies on consumption
behaviour suggests that consumers systematically under- or overspend
based on the nominal (face) value of the foreign currency. However,
previous research is undecided on which direction this effect has, with
the two most prominent models leading to different and mutually
exclusive outcomes. This paper considers the different models and tests
their predictive and explanatory accuracy using a study. It finds that
individuals tend to underspend when the foreign currency is less
numerous than (a fraction of) the home currency and overspend when the
opposite is the case. The paper discusses which model is most accurate
at predicting and explaining this effect and considers limitations as
well as opportunities for future research.
\end{abstract}

@@

Experience suggests that we treat foreign currencies differently to our
home currency and have difficulty adjusting to them. Previous research,
such as the ``money illusion'' effect (\textbf{cite?}) suggests that
individuals are biased towards the nominal (or face value) of prices.
This bias has also been explored in the context foreign currencies, with
research finding systematic effects that arise based on the numerousity
of values in a foreign currency compared to individuals' home
currencies. However, previous research is split on how and why this
effect occurs.

Raghubir and Srivastava (Raghubir and Srivastava 2002) examined the
effect of using foreign currencies on consumption behaviour and argue
that there is a systematic difference in spending behaviour based on the
exchange rate of the foreign currency to an individuals' home currency.
Specifically, they showed that when a foreign currency is a fraction of
the home currency (e.g.~1 unit in home currency is 0.2 units in foreign
currency) consumers tend to overspend in real terms relative to their
home currency. On the other hand, when the foreign currency is a
multiple of the home currency (e.g.~1 unit in home currency is 5 units
in foreign currency) consumers tend to underspend in real
terms.\footnote{For consistency, I will use the terms fraction and
  multiple currencies throughout this paper as defined here.} Raghubir
and Srivastava argue that this effect arises because individuals are
anchored on the face value of the foreign currency which influences the
exchange rate calculations from and to their home currencies. Hence,
Raghubir and Srivastava term the result of this inadequate adjustment
the ``face value effect''.

Wertenbroch, Soman and Chattopadhyay (Wertenbroch, Soman, and
Chattopadhyay 2007) consider the findings by Raghubir and Srivastava and
reiterate their conclusion that individuals inadequately calculate the
values of goods between differnet currencies, leading to systematic
differences in spending behaviour based on the exchange rate between the
foreign and home currencies. However, Wertenbroch, Soman and
Chattopadhyay argue that the process of adjustment occurs differently,
namely that individuals calculate their valuations of goods based on
salient reference points, such as budget constraints. Contrary to
standard theory, Wertenbroch, Soman and Chattopadhyay argue that
consumers care about the difference between a valuation and their budget
rather than the ratio of the two. In the context of foreign currencies,
the nominal value of this difference will diverge, leading to systematic
differences in spending behaviour. However, the direction of this bias
is the exact opposite of the one shown by Raghubir and Srivastava.
Specifically, Wertenbroch, Soman and Chattopadhyay show that consumers
underspend when the foreign currency is a fraction of the home currency
and overspend in the opposite case.

Fang (Fang 2019) further builds on top of the research by Raghubir and
Srivastava and aims to extend it to the domain of ``virtual''
currencies, which refers to non-fiat currencies that are digital in
nature and not controlled or issued by a central bank. Examples of
virtual currencies include video game (or in-game) currencies, rewards
points, airmiles, cryptocurrencies and more to the extent that these are
at least somewhat liquid. Investigating in-game currencies specifically,
Fang shows that consumer valuations systematically deviate when using
these currencies relative to the home currency. Fang's findings
correspond with those of Raghubir and Srivastava and the face value
effect.

This paper aims to build on the aforementioned research in several
important ways. Foremost, previous research is conclusive that there is
an effect that using foreign currencies has on individuals' consumption
behaviour relative to using their home currencies, but is divided on the
direction of this effect. Hence, this paper attempts to evaluate the
different models posed and use a study to test which model most
adequately predicts the perceived behaviour. Second, previous research
has only assigned minor significance to using incentive-compatible
mechanisms to elicit true preferences of participants. These mechanisms
have been established to elicit more accurate responses from
participants that are closer to their true preferences (\textbf{cite?}).
Hence, this study will use the Becker-DeGroot-Marshak (BDM) mechanism to
elicit participants' willingness to pay (WTP).{[}\^{} The BDM mechanism
will be defined and explained in more detail in the Methodology section
below.{]} Raghubir and Srivastava (Raghubir and Srivastava 2002)
expressed the concern that ``despite attempts to make the tasks as
realistic as possible, the studies reported were laboratory experiments,
and issues of generalizability when real money is on the line do arise.
For instance, when people actually exchange foreign currency they may
feel richer or poorer depending on the exchange rate, and the
differences in perceptions of wealth may affect product valuation''.
Similarly, Fang (Fang 2019) noted that implementing the BDM mechanism
was infeasible for the study he conducted but that future research
should focus on using something like it to elicit WTP. Finally, this
paper aims to extend the previous literature on the effects of using
virtual currencies specifically, due to their increasing rise in
importance when transacting or making purchasing decisions online, and
determine whether spending behaviour differs when using them.

First, I present the theoretical frameworks that have been outlined by
previous research in a standard form that makes it easy to compare and
contrast them. These frameworks are the standard model described in
consumer choice theory, the model suggested by Raghubir and Srivastava
and the model suggested by Wertenbroch, Soman and Chattopadhyay. Then, I
consider what each of these frameworks would predict in a concrete case
and show that all three models come to different and mutually exclusive
conclusions. In order to test which of these frameworks is most accurate
in predicting observed behaviour, I conduct a study in which two groups
each formulate their WTP for a specific good in an incentive compatible
way, with the only difference between the groups being the exchange rate
of the currency used to their home currency and, as a result, the
numerousity of the values of their budget and bid. Analysing the study,
I find that consumers tend to underspend when the foreign currency is a
fraction and overspend when the foreign currency is a multiple, which is
predicted by the model proposed by Wertenbroch, Soman and Chattopadhyay
(Wertenbroch, Soman, and Chattopadhyay 2007). However, due to the
limited sample size of the study, this result is not statistically
significant and does not have significant statistical power. In the
discussion, I evaluate the implications of the findings for the
theoretical frameworks outlined, consider alternative explanations for
the perceived effect and argue that the model proposed by Wertenbroch,
Soman and Chattopadhyay is the most accurate in predicting and
describing consumer behaviour when using foreign currencies. Finally, I
will consider the limitations of this paper and opportunities for
further research.

@@

Standard consumer choice theory suggests that individuals formulate
valuations of goods in relation to reference points, such as their
budget (\textbf{cite?}). Specifically, standard theory poses that
consumers evaluate their valuation for a good as a ratio to their budget
in order to determine what proportion of their spending power is
required to purchase the good. To formulate valuations of goods in a
different currency, consumers would first determine the value of the
good in their home currency and subsequently apply the exchange rate to
get their valuation in the foreign currency. Hence, the model follows a
two step process:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Formulate valuation of good as ratio to budget in home currency
\item
  Convert the valuation into foreign currency
\end{enumerate}

@@

where V is the valuation of the good in the foreign currency, \(V_n\) is
the valuation of the good in the home currency (nominal), \(V_r\) is the
valuation of the good in the foreign currency, \(M_n\) is the budget in
the home currency, X is the exchange rate and \(f(x)\) is some function
that uses a valuation relative to the budget in order to compute
\(V_n\). Note that \(V_r\) is simply \(V_n\) multiplied by the exchange
rate, whereas V need not always be identical to \(V_r\).

One assumption that is implicit in this model is that a consumer will
always correctly adjust their valuation ratio based on the exact
exchange rate. However, this might not always be the most rational thing
to do. For example, when exchange rates fluctuate it might not be worth
the effort of a consumer to always use the most up-to-date exchange
rate, but rather to use some approximate value. Further, when an
individual needs to do these exchange rate calculations in their head,
it might not be worth the mental effort to calculate the amount in the
foreign currency perfectly, but an approximation might suffice. For
example, it would be rational to multiply an amount in home currency by
1.2 if the real exchange rate is 1.2113248 and the effort required is
significant, such as when attempting to calculate it in one's hand. By
loosening these assumptions, the standard model is able to account for
small deviations from a consumer's real valuation of a good in a foreign
currency.

@@

Raghubir and Srivastava (Raghubir and Srivastava 2002) put forth a
different model for how consumers formulate their valuation of a good in
a foreign currency based on their studies conducted. Specifically, they
pose that a consumer will first formulate a valuation of a good in their
home currency, before converting that amount into the foreign currency
using the exchange rate. However, they argue that this calculation is
biased by the nominal value of the good in the home currency (in cases
where a consumer does not determine their valuation but observes a
price, this calculation would be biased by the nominal value of the good
in the foreign currency). This model may be formulated as follows:

@@

where α is a weighting parameter such that \(α ∈[0,1]\). Importantly,
Raghubir and Srivastava (Raghubir and Srivastava 2002) argue that
\(α > 0\) and that V is hence not equal to \(V_r\) but is biased towards
\(V_n\).

@@

Wertenbroch, Soman and Chattopadhyay (Wertenbroch, Soman, and
Chattopadhyay 2007) build on top of Raghubir and Srivastava, but put
forth a different model that they argue is more representative of the
behaviour of actual consumers. Specifically, they agree with the
standard model that valuations are made in relation to salient reference
points, such as a consumers' budget, and argue that the model put forth
by Raghubir and Srivastava fails to account for this. However, they
disagree with the standard model that valuations use a ratio of price to
budget, but instead argue that consumers think in terms of differences.
Hence, instead of considering what fraction of their purchasing power is
required to buy a good, consumers consider how much of their budget is
left over. However, Wertenbroch, Soman and Chattopadhyay argue that
consumers inadequately adjust this difference to the exchange rate and
are thus biased when considering goods in different currencies since the
nominal differences between price and budget will differ when
considering different currencies. This model can be formulated as
follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Formulate valuation of good as difference to budget in home currency
\item
  Convert the valuation into foreign currency
\item
  Adjust using the difference to budget in the foreign currency
\end{enumerate}

@@

where \(M_r\) is the budget in the foreign currency (real). Importantly,
Wertenbroch, Soman and Chattopadhyay (Wertenbroch, Soman, and
Chattopadhyay 2007) argue that \(α > 0\) and that V is hence not equal
to \(V_r\) but is biased towards \(M_r - V_r\).

@@

@@

The study itself was a survey split into four parts: consent,
introduction, auction and follow-up Questions. The entire experiment
instructions can be found in the Experiment Instructions section of the
Appendix. The first section asked participants to read through and sign
the consent form in order to ensure that they were informed about the
study and what data would be collected. In the introductory section, the
BDM mechanism was explained to participants and they were asked to
complete four question that tested their understanding of the mechanism.
The aim of this section was to ensure that participants properly
understood how the mechanism worked in order to elicit their true WTP
during the main part of the study.

The auction was the main part of the study, with the aim of eliciting
the participants' WTP on a specific item and comparing these across
different exchange rates used. In this section, participants were
randomly split into two groups with almost-identical instructions. The
participants were instructed to bid on an item shown below, a tabletop
airhockey table, using the virtual currency ``BlueCoin''. The auction
was explained to follow the BDM mechanism and it was also stated that
three participants of the experiment would be randomly selected for
their auction to be simulated and to receive the outcome (either their
budget in Pounds or the item and remaining budget). The only difference
between the groups was the exchange rate between BlueCoin and the
British Pound (£) and the endowment that the participants received in
BlueCoin. The first group received an endowment of 10 BlueCoin where 1
BlueCoin equals £5 and the second group received an endowment of 250
BlueCoin where 1 BlueCoin equals £0.2. To complete the auction,
participants were asked to respond with their bid, in BlueCoin.

@@

Since the aim of the study was to determine whether there were any
systematic differences in WTP between consumers when the only differing
factor is the numerousity of the currency used, I decided to not include
a control group that bid on the item using Pounds, which was the design
of some of the experiments run by Raghubir and Srivastava (Raghubir and
Srivastava 2002). This allowed for a simpler design and more power given
the same sample size. Further, unlike the studies run by Raghubir and
Srivastava (Raghubir and Srivastava 2002) and Wertenbroch, Soman and
Chattopadhyay (Wertenbroch, Soman, and Chattopadhyay 2007), I did not
use a fiat currency or mix of fiat and virtual currencies, but used only
one virtual currency for both groups. On the one hand, this
simplification was due to a focus of wether the effects described in
previous papers also held in similar ways for virtual currencies. On the
other hand, it also allowed me to preclude any biases or previous
experience that participants might have had towards or with certain
currencies. Raghubir and Srivastava (Raghubir and Srivastava 2002)
acknowledge that ``the face value effect is due to the accessibility and
perceptual salience of the face value of the foreign currency \ldots{}
{[}which{]} is likely to depend on the extent to which an individual has
the opportunity or the time available to process exchange rate
information and/or has experience in using a particular foreign
currency''. Further, Alter and Oppenheimer (Alter and Oppenheimer 2008)
show that people use ``familiarity and fluency'' when valuing goods in
different currencies. To bias participants the least, I chose the
relatively nondescript name for a fictional, virtual currency: BlueCoin.

Another important decision in the design of the experiment was which
item to select that a majority of participants would have a non-zero
WTP. This is important since if the majority or even the entirety of
participants were to bid zero on the item, then I would only learn that
this was the participants' WTP for that good, but not whether there are
any systematic differences in the WTP when using different currencies.
Hence, I selected a tabletop airhockey table, something that few enough
people have so that they would not not want another one, but enough
people want with a non-zero WTP, especially among a student population.
This turned out to be a good choice, since only six participants of the
experiment stated a WTP of 0.

In designing the main part of the experiment, I chose to conduct an
auction, specifically using the BDM mechanism, in order to have an
incentive-compatible method of eliciting the participants' WTP. The
BDM's incentive compatibility in eliciting WTP is a well-established
phenomenon that is superior to using participants' stated preferences
(\textbf{cite?}). The BDM is essentially a sealed-bid, second price
auction that is, in its' most common variant, played against a computer
drawing a bid from a known distribution. If the players' bid is greater
than or equal to the computers' bid, they win and pay the computers'
bid, otherwise they lose. As mentioned above, the studies conducted by
Raghubir and Srivastava (Raghubir and Srivastava 2002), Wertenbroch,
Soman and Chattopadhyay (Wertenbroch, Soman, and Chattopadhyay 2007)
(with the exception of one permutation of one of the studies) and Fang
(Fang 2019) did not implement incentive-compatible methods of eliciting
the participants' preferences, making this experiment an important
addition to these previous studies in examining whether the effects
described still hold when a more truthful way of eliciting WTP is used.
In order to incentivise the participants to participate, I ran a lottery
that paid out three randomly-chosen participants based on their actual
choices made in the auction and by simulating the BDM mechanism using a
random number generator. Due to budget constraints, I was only able to
play this lottery for three people, but given that the sample size of
the experiment was relatively small, the potential to receive one's
outcome is likely to have helped in eliciting the participants' true
WTP.

Finally, given the budget, I was able to give participants an endowment
of £50 each, which was around double of what the item costs to buy.
Assuming that the item is priced roughly around the average WTP (which
turned out to be correct), this endowment was chosen to leave enough
room for people with higher WTP to make their bid and to capture this in
the data. This is important for the goals of the experiment, since if I
had, for example, chosen an endowment of £20, the majority of
participants might have bid their maximum amount, hindering my ability
to meaningfully compare WTP across the two groups. I chose relatively
simple exchange rates of 5 and \(1/5\) so that the exchange rate
calculations would be relatively easy to make, allowing me to be more
confident that differences between groups were not due to the mental
effort required to do calculations but due to something else. Raghubir
and Srivastava (Raghubir and Srivastava 2002) acknowledge that ``the
reliance on face value may be a function of the ease with which the
foreign money can be converted. \ldots{} Future research should examine
the asymmetric and nonlinear nature of this effect''. Since the goal of
this study is to investigate that an effect exists that is not solely
based on a high difficulty or large mental effort required when doing
specific, difficult calculations, I aimed to make the calculations as
easy as possible in order to isolate the effect of the exchange rate and
nominal value of the foreign currency. Further, since the study was
conducted online, participants were easily able to and allowed to use
calculators for any part.

@@

In setting up the experiment, I incorrectly used the BDM mechanism.
Instead of explaining it as a sealed-bid second-price auction, the
experiment instructions explained it as a sealed-bid first-price
auction. Unlike in a second-price auction, the participants of a first
price auction do not have a (weakly) dominant strategy to bid their true
valuations, in fact there generally is no dominant strategy in such an
auction (\textbf{cite?}). This is true even in this auction in which
participants knew that they were bidding against a computer agent who
would randomly draw a bid from a uniform distribution from 0-50, with an
expected bid of 25. Indeed, the optimal strategy by participants was to
choose their bid b by maximising their payoff p:

\[\frac{b}{50} * u(x = 1, 50 - b) + (1 - \frac{b}{50}) * u(x = 0, 50)\]
